Prefix-free graphs and suffix array construction in sublinear space

### 2023-06-21 manuscript/main.tex
added path_join

\subsection{Iterator preparation}
In order to prepare the prefix-free graph to generate the suffix array of
original pangenome we need to create several data structures.
First, we concatenate all the prefix-free segments into a single string using
a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.

Next, for the segment join we calculate the suffix array and the LCP array.
For this we can use the linear time, linear space algorithms mentioned previously,
since the segment join is assumed to be much smaller than the original pangenome.

Next, for each suffix of a segment join, we need to calculate the corresponding
segment ID and the segment position.
These two arrays represent the current suffix of a segment join starts with a
particular segment and the position in this segment.
This can be computed using an inverse permutation of a suffix array of a segment
join in linear time.
To illustrate the procedure, consider the segment join of our running example in
Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in current
segment by linearly iterating the segment join and incrementing ID and position
accordingly.
Then, applying the inverse permutation of a suffix array to these arrays will
change the order of computed values in a way corresponding to the sorted suffixes.
The Table \ref{tab:suffix} shows the resulting suffix array, LCP array, segment ID array and segment position array.

@@

In the Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
In order to be able to reconstruct the suffix array, we need to store some
additional information for each segment.
For each segment, we store it's length, it's starting positions in the implicit
concatenated pangenome and the rank of it's right context.

@@

>>>>

\subsection{Iterator preparation}
In order to prepare the prefix-free graph to generate the suffix array of
original pangenome we need to create several data structures.
First, we concatenate all the prefix-free segments into a single string using
a separator $\#$ and append a sentinel $\$$.
We will call this concatenation \emph{segment join}.
An example of a segment join is in the Figure \ref{fig:ids_and_positions}.

Next, for the segment join we calculate the suffix array and the LCP array.
For this we can use the linear time, linear space algorithms mentioned previously,
since the segment join is usually much smaller than the original pangenome.

Next, for each suffix of a segment join, we need to calculate the corresponding
segment ID and the segment position.
These two arrays represent the current suffix of a segment join starts with a
particular segment and the position in this segment.
This can be computed using an inverse permutation of a suffix array of a segment
join in linear time.
To illustrate the procedure, consider the segment join of our running example from
Figure \ref{fig:ids_and_positions}.
Each position of the join can be assigned a segment ID and a position in current
segment by linearly iterating the segment join and incrementing ID and position
accordingly.
Then, applying the inverse permutation of a suffix array to these arrays will
change the order of computed values in a way corresponding to the sorted suffixes.
The Table \ref{tab:suffix} shows the resulting suffix array, LCP array,
segment ID array and segment position array.

@@

In the Table \ref{tab:suffix}, one row can represent multiple positions of the
pangenome.
To identify these positions, we store some additional information in a second
table \ref{tab:segment}.
For each segment, we store it's length, it's starting positions in the pangenome
and the rank of the right context of these positions.
To calculate the starting positions and the ranks of the right contexts we use
a path join.
Similarly to segment join, a path join is a concatenation with delimiters $\#$
and sentinel $\$$, but now constructed by concatenating the paths.
An example of a path join for our running example is in the Figure \ref{fig:path_join}.
Starting positions can then be calculated by cumulatively summing the lengths
of the segments in path join and subtracting the overlaps.
Computation of the ranks is more involved.
We need to construct the suffix array of the path join and find it's inverse
permutation $ISA$.
This gives us ranks for each position in the path join.
To determine the rank of the right context for position $i$, we simply take
the value of the ranks $ISA[i+1]$.
Finaly, we store the starts and ranks sorted by the rank values in the segment
table as shown in Table \ref{tab:segment}.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{images/path_join.png}
    \caption{}
    \label{fig:path_join}
\end{figure}

@@

\subsection{Iterating suffix array}
With the previous tables in memory, we have all we need to generate the suffix
array entries.
Due to the prefix-freeness of the segments and the construction of the segment join
we can ignore all characters after the first delimiter of $\texttt{suffix}[i..]$.
Then, each row in the suffix table represents a suffix of a particular segment.

There are five cases we need to consider.
\begin{itemize}
    \item suffix represents a separator or the sentinel
    \item suffix represents a trigger or it's suffix
    \item suffix longer than the trigger representing a single position in the pangenome
    \item suffix longer than the trigger representing several positions in the pangenome
\end{itemize}

We can skip the first $n+1$ rows of the suffix table, because these suffixes
do not yield any suffix array entry.
